Using spatially biased variables in ecosystem condition accounting with a GIS based workflow

### 2024-01-29 manuscript/HIAs.qmd
make variables continous

library(tidyverse)
library(knitr)
library(sf)
library(tmap)
library(stars)
library(terra)
library(tidyterra)
library(ggtext)
library(cowplot)

@@

@@

<!--# Need to update this figure -->

@@

| Variable code        | Variable name                                    | Measurement unit                                                                        | Description                                                                                                    | Reference                          |
|-------------|-------------|---------------|--------------------|-------------|
| 7FK                  | Prevalence of alien species                      | Unitless, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                             | @runehalvorsen2019                 |
| 7SE                  | Human caused abration or abration-caused erosion | Unitless, ordinal, non-linear scale from 1 to 4.                                        | Measures the frequency of imagined 4 m^2^ quadrats layed over the area that has some sign of abration          | @runehalvorsen2019                 |
| PRSL                 | *as above*                                       | Unitless, ordinal, non-linear scale from 0 to 7.                                        | Same as 7SE, but recorded at a higher resolution                                                               | @miljødirektoratet                 |
| 7TK                  | Tracks from large vehicles                       | Unitless, ordinal, non-linear scale from 1 to 4.                                        | Measures the frequency of imagined 100 m^2^ quadrats layed over the area that has some signs of vehicle tracks | @rune_halvorsen_dokumentasjon_2019 |
| PRTK                 | *as above*                                       | Unitless, ordinal, non-linear scale from 0 to 7.                                        | Same as 7TK, but recorded at a higher resolution                                                               | @miljødirektoratet                 |
| Infrastructure Index | Infrastructure Index                             | Unitelss linear scale from 0 to 13.2                                                    | Unitelss index ranging from from 0 to 13.2                                                                     | @erikstad_index_2023               |
| HIA                  | Human Impact area                                | Ordinal, non-linear scale from 1 to 4                                                   | A categorical representation of the infrastructure index                                                       | this paper                         |

>>>>

library(tidyverse)
library(knitr)
library(sf)
library(tmap)
library(stars)
library(terra)
library(tidyterra)
library(ggtext)
library(cowplot)
library(units)

@@

```{r convertToPercent}
#| eval: true

# I now want to take the variable and normalise it before I can then combine them
# despite them being on different scales
# I will first normalise by converting into %.
# Remeber the ordinal categories represnts frequency ranges

# The data is strongly right skewed, so simply taking the center value of each
# bin will not work.
naturetypes %>% 
  ggplot()+
  theme_bw() +
  geom_histogram(aes(x = value),
                 binwidth = 1) +
  facet_wrap(.~variable,
             scales = "free")

# I will use th lower bound for each bin.
# The exception in when the variable is 1, because then the lower bound
# is 0, same as when the variable is 0.
# For these I will set manually a slightly higher value.

naturetypes <- naturetypes %>%
  mutate(value = case_when(
    # selecting the variable that follow the same 4 step scale
    variable %in% c("7TK", "7SE", "7FA") ~ 
      case_match(value,
             0 ~ 0,
             1 ~ mean(c(0, 1/16))*100,
             2 ~ 1/16*100,
             3 ~ 50), # note that it is not possible to get a value of 1
    # selecting the eight step variables
    variable %in% c('PRTK', 'PRSL') ~ 
      case_match(value,
             0 ~ 0,
             1 ~ 1.5,
             2 ~ 3,
             3 ~ 6.25,
             4 ~ 12.5,
             5 ~ 25,
             6 ~ 50,
             7 ~ 75)
    )
  )

naturetypes %>% 
  ggplot()+
  theme_bw() +
  geom_histogram(aes(x = value),
                 binwidth = 1,
                 color = "orange",
                 fill = "orange") +
  facet_wrap(.~variable,
             scales = "free")

```{r}
#| eval: true
# Now I make the data wide, and remove 7TK and 7SE if PRTK or PRSL are present,
# respectively

naturetypes_wide <- naturetypes |>
  # Column names starting with a number is problematic, so adding a prefix
  mutate(variable = paste0("var_", variable)) |>
  pivot_wider(
    names_from = "variable",
    values_from = "value",
    id_cols = "id") |>
    as_tibble()

head(naturetypes_wide, 10)
```

```{r}
#| eval: true
# First I will combine 7TK and PRTK, and also 7SE and PRSL.

naturetypes_wide <- naturetypes_wide %>%
  mutate(TK = if_else(
    is.na(var_PRTK), var_7TK, var_PRTK),
    SE = if_else(
      is.na(var_PRSL), var_7SE, var_PRSL)
    )

plot_grid(
naturetypes_wide %>%
  as_tibble() |>
  count(SE,
        name = "sum") |>
  ggplot(
    aes(
      x = factor(SE),
      y = sum))+
  geom_bar(stat="identity",
           fill="grey",
           colour = "black")+
  theme_bw(base_size = 12)+
  labs(x = "7SE or PRSL score",
       y = "Number of localities"),

naturetypes_wide %>%
  as_tibble() |>
  count(TK,
        name = "sum") |>
  ggplot(
    aes(
      x = factor(TK),
      y = sum))+
  geom_bar(stat="identity",
           fill="grey",
           colour = "black")+
  theme_bw(base_size = 12)+
  labs(x = "7TK or PRTK score",
       y = "Number of localities")
)

# The NA's represnets localities where just one of the two variables 
# (then thinking 7SE and PRSL as the same variable)
# is recorded. To combine these into one metric ADSV, I could take the 
# one with the highest value (worst-rule) og the sum.
# Sum is problematic is not all locations ahve two values to sum together.
# But the other option is problematic since I think field workers often
# tend to split the effects over two variables is they have that option.
# And if we have 50% vehicle damaeg and 50% hiking damage, that is no doubt
# worst than just having 50% of either. So I will use the sum, despite its issues.
```

```{r}
#| eval: true

# Taking he sum of 7SE and 7TK (incl the PR.. variabels)
naturetypes_wide <- naturetypes_wide |>
  rowwise() |>
  mutate(ADSV = sum(c(SE, TK), na.rm=TRUE))

naturetypes_wide %>%
  as_tibble() |>
  count(ADSV,
        name = "sum") |>
  ggplot(
    aes(
      x = (ADSV),
      y = sum))+
  geom_bar(stat="identity",
           fill="grey",
           colour = "black")+
  theme_bw(base_size = 12)+
  labs(x = "Summed ADVS score",
       y = "Number of localities") +
    scale_x_continuous(
    labels = scales::label_number(accuracy = 1))

```{r}
#| eval: true
# Now I will copy these ADVS-values into the sf object again
naturetypes <- naturetypes |>
  left_join(naturetypes_wide |> select(id, var_7FA, ADSV), by = "id")
```

@@

@@

| Variable code        | Variable name                                    | Measurement unit                                                                        | Description                                                                                                    | Reference                          |
|--------------|--------------|--------------|----------------|--------------|
| 7FK                  | Prevalence of alien species                      | Unitless, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                             | @runehalvorsen2019                 |
| 7SE                  | Human caused abration or abration-caused erosion | Unitless, ordinal, non-linear scale from 1 to 4.                                        | Measures the frequency of imagined 4 m^2^ quadrats layed over the area that has some sign of abration          | @runehalvorsen2019                 |
| PRSL                 | *as above*                                       | Unitless, ordinal, non-linear scale from 0 to 7.                                        | Same as 7SE, but recorded at a higher resolution                                                               | @miljødirektoratet                 |
| 7TK                  | Tracks from large vehicles                       | Unitless, ordinal, non-linear scale from 1 to 4.                                        | Measures the frequency of imagined 100 m^2^ quadrats layed over the area that has some signs of vehicle tracks | @rune_halvorsen_dokumentasjon_2019 |
| PRTK                 | *as above*                                       | Unitless, ordinal, non-linear scale from 0 to 7.                                        | Same as 7TK, but recorded at a higher resolution                                                               | @miljødirektoratet                 |
| Infrastructure Index | Infrastructure Index                             | Unitelss linear scale from 0 to 13.2                                                    | Unitelss index ranging from from 0 to 13.2                                                                     | @erikstad_index_2023               |
| HIA                  | Human Impact area                                | Ordinal, non-linear scale from 1 to 4                                                   | A categorical representation of the infrastructure index                                                       | this paper                         |
